---
title: "ERSST"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ersst

Provides for download, archiving and access to [Extended Reconstructed Sea Surface Temperature 'ersst'](https://www.ncei.noaa.gov/products/extended-reconstructed-sst) online and local datasets.

### [Citation](https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00927)

```
Boyin Huang, Peter W. Thorne, Viva F. Banzon, Tim Boyer, Gennady Chepurin, Jay H. Lawrimore, Matthew J. Menne, Thomas M. Smith, Russell S. Vose, and Huai-Min Zhang (2017): NOAA Extended Reconstructed Sea Surface Temperature (ERSST), Version 5. NOAA National Centers for Environmental Information. doi:10.7289/V5T72FNM [access date, monthly from 2021-04-15].
```

### Requirements

From CRAN...

  + [R v4+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [readr](https://CRAN.R-project.org/package=readr)
  + [terra](https://CRAN.R-project.org/package=terra)
  + [xml2](https://CRAN.R-project.org/package=xml2)
  + [httr](https://CRAN.R-project.org/package=httr)
  + [rvest](https://CRAN.R-project.org/package=rvest)
  + [stars](https://CRAN.R-project.org/package=stars)
  + [PCICt](https://CRAN.R-project.org/package=PCICt)

### Installation

```
remotes::install_github("BigelowLab/ersst")
```

### Usage

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(sf)
  library(ersst)
  library(stars)
})

```

#### Working with points.  

We provide a small example of points from locations associated with buoys off of the US southeast coast.

```{r}
# read in example South Atlantic Bight points
x <- read_sab()

# generate a ersst url for a given date
url <- ersst_url("2018-12-18")

# download
temp_file <- tempfile(fileext = ".nc")
ok <- download.file(url, temp_file)

# open the resource
X <- ncdf4::nc_open(temp_file)

# extract the data 
covars <- ersst::extract(x, X, varname = ersst_vars(X))

# bind to the input
(y <- dplyr::bind_cols(x, covars))
```

#### Working with bounding boxes (from points or polygons). 

Learn more about working with [stars](https://CRAN.R-project.org/package=stars) objects in the [vignettes](https://r-spatial.github.io/stars/).  Here we drop time and depth as the ERSST data has only one depth and each file holds one time.

```{r}
# derive the bounding box of the points and convert to sf coordinates
bb = sf::st_bbox(x) |>
  sf::st_as_sfc()

(covars <- ersst::extract(bb, X, varnames = ersst::ersst_vars(X)))
```

Now let's see what it looks like on a quick map.

```{r}
plot(covars, attr = 'sst', axes = TRUE, reset = FALSE)
plot(sf::st_geometry(x), add = TRUE, col = "orange", 
     pch = 19, cex = 2)
```


```{r}
# cleanup
ncdf4::nc_close(X)
```

